Two Decades of Smalltalk VM Development

### 2018-08-17 paper.tex
minor fixes

\paragraph{Templates.} The JIT generates a sequence of machine instructions for a given pattern of bytecodes. We use the term template to describe the sequence of machine instructions generated for a pattern of bytecodes, even though we do not like to reduce the JIT to a simple template-based JIT since each bytecode generates slightly different machine instructions based on register pressure, a simulated stack and a few heuristics. Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates. Since we also have a bytecode assembler that allows us to manufacture bytecoded methods we can in practice exercise any part of the JIT by creating a suitable bytecode method. Although for methods of the complexity produced by the adaptive optimiser this is hardly pleasurable.

@@

In this section we describe our personal experience using the simulation infrastructure to develop the VM. The different subsection consists in different tasks we had to do, emphasising how we used the simulator and how it helped us perform the each given task.

@@

In the recent years, we added support for a more aggressive JIT with speculative optimizations through a bytecode to bytecode optimizer \cite{SistaFaststart-up}, re-using the existing template JIT as a back-end. To be able to generate efficient bytecodes in the bytecode to bytecode optimizer, we had to introduce new unsafe bytecodes allowing, for example, accesses in arrays without any checks (type and bounds checks) \cite{SistaBytecodeSet}. For each new bytecode, we introduced new templates in the existing JIT to generate efficient machine code for the optimized methods using the new bytecodes. Once all the basic unit tests worked, we ran the VM with the speculative optimizer, which executed optimized code, and got a crash. We could figure out which bytecode method was triggering the crash, but we had no idea from which template the crash came from. In addition, optimized methods include many inlined methods, making them very large, so it was difficult to figure out where the issue comes from just by looking at thousands of bytecodes.

>>>>

\paragraph{Templates.} The JIT generates a sequence of machine instructions for a given pattern of bytecodes. We use the term template to describe the sequence of machine instructions generated for a pattern of bytecodes, even though we do not like to reduce the JIT to a simple template-based JIT since each bytecode generates slightly different machine instructions based on register pressure, a simulated stack and a few heuristics. Since the JIT is template-based, in-image compilation is very convenient to develop and optimize each of the JIT templates.

Since we also have a bytecode assembler that allows us to manufacture bytecoded methods we can in practice exercise any part of the JIT by creating a suitable bytecode method. Although for methods of the complexity produced by the adaptive optimiser this is hardly pleasurable.

@@

In this section we describe our personal experiences using the simulation infrastructure to develop the VM. The different subsections consist in different tasks we had to implement and in each case we emphasise how we used the simulator and how it helped us perform the each given task.

@@

In the recent years, we added support for a more aggressive JIT with speculative optimizations through a bytecode to bytecode optimizer \cite{SistaFastStartup}, re-using the existing template JIT as a back-end. To be able to generate efficient bytecodes in the bytecode to bytecode optimizer, we had to introduce new unsafe bytecodes allowing, for example, accesses in arrays without any checks (type and bounds checks) \cite{SistaBytecodeSet}. For each new bytecode, we introduced new templates in the existing JIT to generate efficient machine code for the optimized methods using the new bytecodes. Once all the basic unit tests worked, we ran the VM with the speculative optimizer, which executed optimized code, and got a crash. We could figure out which bytecode method was triggering the crash, but we had no idea from which template the crash came from. In addition, optimized methods include many inlined methods, making them very large, so it was difficult to figure out where the issue comes from just by looking at thousands of bytecodes.
